\bf An Examination of Sport Climbing Competition Format and Scoring System

### 2021-07-30 manuscript/Rmd/draft.Rmd
Add bib file
Bed time

abstract: |
  The purpose of this paper is to investigate sport climbing, one of the sports making its debut on the Olympics stage at Tokyo 2020. In particular, we take a closer look at the controversial competition format and scoring system of sport climbing... Simulation. Data Analysis: drop and re-rank, correlations
# bibliography: bib_example.bib
output: rticles::asa_article
---

@@

In 2016, the International Olympic Committee announced the addition of five new sports to the 2020 Summer Olympics in Tokyo, Japan, which would then reschedule for 2021 due to the impact of the COVID-19 global pandemic. The five new features to Tokyo 2020's competitions program include baseball/softball, karate, skateboard, sports climbing, and surfing. One of the new sports, sport climbing, got our attention, specifically because of its unique scoring system and the fact that only one set of medals is awarded for each gender.

@@

At the 2020 Summer Olympics, both men's and women's sport climbing competitions begin with 20 climbers who has previously qualified for the Olympics from qualifying events held in 2019 and 2020. All 20 athletes compete in each of the three disciplines in the qualification round, and their performances in each concentration are ranked from 1 to 20.  A competitor's total score is then computed as the product of their ranks in the three events and the lower product is better; specifically, 
\begin{equation}
Score_i = R^S_i\times R^B_i\times R^L_i,
\end{equation} 
where $R^S_i$, $R^B_i$, and $R^L_i$ are the ranks of the $i$-th competitor in speed climbing, bouldering, and lead climbing, respectively.

The top 8 finishers in the qualification round advance to the finals where they once again compete in all three of speed climbing, bouldering, and lead climbing. The total score for each climber in the final stage is determined by multiplying their ranks in each discipline, similar to the qualification round. This implies that the climber with the lowest product of ranks in the final wins the gold medal. This type of scoring system heavily rewards high finishes and relatively ignores poor finishes.  For instance, if climber A finished 1st, 20th, and 20th and climber B finished 10th, 10th, and 10th, climber B would have a score of 1000 whereas climber A would have a much better score of 400, despite finishing last in 2 out of 3 of the events.

Heavily criticized

@@

The manuscript is outlined as follows. We first begin with some descriptions of the data and methods in Section 2. Our analyses and results are then presented in Sections 3 and 4. Finally, in Section 5, we summarize up our main findings and provide a discussion to close out the paper.

@@

Our first sets of data come from a simulation study that we conducted, with the purpose of examining the rankings and scoring for climbers in both qualification and final rounds. For each round, we performed 10000 simulations, and this was accomplished by randomly assigning the ranks of each event to every participant, with the assumption that the ranks are uniformly distributed. After the completion of the simulations, we calculated the total scores for every simulated round, as well as the final standings for the climbing athletes. The simulation results allow to answer questions about various topics, including the distributions of scores for qualifying and final rounds, and the probabilities of advancing to the finals or winning a medal, given certain conditions.

Table 1 and Figure 1 are numerical and visual summaries of climbing total score obtained from our simulated data for the qualification and final rounds.

```{r}
library(tidyverse)
library(knitr)
library(xtable)
theme_set(theme_minimal())
```

@@

@@

For our analysis on the relationship between the rankings of the events and the final result, we used data from the 2018 Youth Olympics Women's Qualification. Figure 2 is a scatterplot and correlation matrix between the ranks of the individual events and the final standings, with Kendall’s Tau (Kendall Rank Correlation Coefficient) as our measure of ordinal association between the quantities. It is evidently clear that there exists a strong and positive correlation between the ranks of bouldering and lead climbing, and as a results, the standings of these two events are highly correlated with the final rankings. On the other hand, the correlation with the final rank is not as strong for speed climbing. Thus, speed climbers are facing a huge disadvantage in this scoring system, compared to those that are specialized in the other two concentrations. 
<!-- This trend also holds for most of the past competitions. -->

```{r}
wq <- read_csv("https://raw.githubusercontent.com/qntkhvn/climbing/main/data/2018_youth_olympics/women_qual.csv")
library(GGally)
```

```{r, fig.cap = "Kendall's rank correlations - 2018 World Championship, Women's Qualification"}
wq %>% 
  select(rank, speed, bould, lead) %>% 
  ggpairs(diag = "blank",
          axisLabels = "internal",
          upper = list(continuous = wrap("cor", method = "kendall", stars = FALSE))) +
  theme_bw() +
  theme(panel.grid.major = element_blank())
```

@@

wf_gg <- drop_rerank(wf) %>% 
  mutate(last = fct_reorder(last, -rank),
         rank = as.factor(rank),
         rank_change = ifelse(
           rank_drop %in% c(0, 1, 4, 6) |
             rank_drop == 2 & rank %in% 1:3 |
              rank_drop == 3 & rank %in% c(1, 4, 5) |
                rank_drop == 5 & rank %in% c(1, 5),
           "no", "yes")) %>%
  ggplot(aes(x = last, y = total, fill = rank_change)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = rank), hjust = -0.2, size = 3) +
  coord_flip() +
  geom_rect(aes(xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf),
            data = ~ filter(., rank_drop %in% c(2, 3, 5)), 
            color = "black", size = 1.5, fill = NA, inherit.aes = FALSE) +
  facet_wrap(~ rank_drop, nrow = 2, ncol = 4) +
  expand_limits(y = 62, x = 0:7) +
  scale_fill_manual(values = c("grey", "chocolate")) +
  labs(y = "\nScore", x = "Women") +
  theme(axis.ticks = element_blank(),
        panel.grid.minor = element_blank())
```

@@

>>>>

abstract: |
  The purpose of this paper is to investigate sport climbing, one of the sports making its debut on the Olympics stage at Tokyo 2020. In particular, we take a closer look at the controversial competition format and scoring system of sport climbing... Simulation. Data Analysis: drop and re-rank, correlations
bibliography: references.bib
output: rticles::asa_article
---

@@

In 2016, the International Olympic Committee announced the addition of five new sports to the 2020 Summer Olympics in Tokyo, Japan, which would then reschedule for 2021 due to the impact of the COVID-19 global pandemic. The five new features to Tokyo 2020's competitions program include baseball/softball, karate, skateboard, sports climbing, and surfing [@ioc2016]. One of the new sports, sport climbing, got our attention, specifically because of its unique scoring system and the fact that only one set of medals is awarded for each gender.

@@

The decision to combine the three climbing events and only award one set of medals for both men's and women's events in the Olympics has received a large amount of criticism from climbing athletes all over the world. In a series of interviews conducted by Climbing Magazine in 2016, a number of climbers shared their thoughts and concerns about the new Olympics climbing format. Climber Lynn Hill compared the idea of combining speed climbing, bouldering, and lead climbing to "asking a middle distance runner to compete in the sprint". She then added "Speed climbing is a sport within our sport". Other climbers also hold the same opinion as Hill regarding speed climbing, using words and phrases like "bogus", "a bummer", "less than ideal", "not in support", and "cheesy and unfair" to describe the new combined competition format. Courtney Woods stated "Speed climbers will have the biggest disadvantage because their realm isn’t based on difficult movements". Mike Doyle believed "Honestly, the people that will suffer the most are the ones that focus only on speed climbing. Those skills/abilities don’t transfer as well to the other disciplines". Most climbers also expressed their hope for a change in the format in future tournaments, with some calling for each discipline to have its own set of medals. See @blanchard2016 for more information on these interviews.

In terms of participation and scoring, both sport climbing competitions for male and female at the 2020 Summer Olympics begin with 20 climbers who has previously qualified for the Olympics from qualifying events held in 2019 and 2020. All 20 athletes compete in each of the three disciplines in the qualification round, and their performances in each concentration are ranked from 1 to 20. A competitor's combined score is computed as the product of their ranks in the three events; specifically, 
\begin{equation}
Score_i = R^S_i\times R^B_i\times R^L_i,
\end{equation} 
where $R^S_i$, $R^B_i$, and $R^L_i$ are the ranks of the $i$-th competitor in speed climbing, bouldering, and lead climbing, respectively.

The top 8 lowest-scoring climbers in the qualification round advance to the finals where they once again compete in all three of speed climbing, bouldering, and lead climbing. The total score for each person in the final stage is determined by multiplying their ranks in each discipline, similar to the qualification round. This implies that the climber with the lowest product of ranks in the final wins the gold medal. This type of scoring system heavily rewards high finishes and relatively ignores poor finishes.  For instance, if climber A finished 1st, 20th, and 20th and climber B finished 10th, 10th, and 10th, climber B would have a score of 1000 whereas climber A would have a much better score of 400, despite finishing last in 2 out of 3 of the events.

@@

https://www.climbing.com/competition/olympics/a-guide-to-the-olympic-climbing-format/

Soccer points: @haugen2008

Cross country scoring: @hammond2007, @boudreau2018

Decathlon scoring: @westera2006

The manuscript is outlined as follows. We first begin with some descriptions of the data and methods in Section \ref{data-and-methods}. Our analyses and results are then presented in Sections \ref{simulations} and \ref{data-analysis}. Finally, in Section \ref{conclusion-and-discussion}, we summarize up our main findings and provide a discussion to close out the paper.

@@

Our first sets of data come from a simulation study that we conducted, with the purpose of examining the rankings and scoring for climbers in both qualification and final rounds. For each round, we performed 10000 simulations, and this was accomplished by randomly assigning the ranks of each event to every participant, with the assumption that the ranks are uniformly distributed. After the completion of the simulations, we calculated the total scores for every simulated round, as well as the final standings for the climbing athletes. The simulation results allow to answer questions about various topics, including the distributions of scores for qualifying and final rounds, and the probabilities of advancing to the finals or winning a medal, given certain conditions. First of all, Table 1 and Figure 1 are numerical and visual summaries of climbing total score obtained from our simulated data for the qualification and final rounds.

```{r}
library(tidyverse)
library(xtable)
theme_set(theme_minimal())
```

@@

Brief paragraph on IIA

@@

For our analysis on the relationship between the rankings of the events and the final result, we used data from the 2018 Youth Olympics Women's Qualification. Figure 2 is a scatterplot and correlation matrix between the ranks of the individual events and the final standings, with Kendall’s Tau [@kendall1938] as our measure of ordinal association between the quantities. It is evidently clear that there exists a strong and positive correlation between the ranks of bouldering and lead climbing, and as a results, the standings of these two events are highly correlated with the final rankings. On the other hand, the correlation with the final rank is not as strong for speed climbing. Thus, speed climbers are facing a huge disadvantage in this scoring system, compared to those that are specialized in the other two concentrations. 
<!-- This trend also holds for most of the past competitions. -->

```{r}
library(GGally)
mq <- read_csv("https://raw.githubusercontent.com/qntkhvn/climbing/main/data/2018_youth_olympics/men_qual.csv")
wq <- read_csv("https://raw.githubusercontent.com/qntkhvn/climbing/main/data/2018_youth_olympics/women_qual.csv")
```

```{r, fig.cap = "Kendall's rank correlations - 2018 World Championship, Men's and Women's Qualifications", fig.height=8, fig.width=6}
mcor <- mq %>% 
  select(Overall = rank, Speed = speed, Bouldering = bould, Lead = lead) %>% 
  ggpairs(diag = "blank",
          axisLabels = "internal",
          upper = list(continuous = wrap("cor", method = "kendall", stars = FALSE))) +
  theme(panel.grid.major = element_blank()) +
  labs(y = "Men\n")

wcor <- wq %>% 
  select(Overall = rank, Speed = speed, Bouldering = bould, Lead = lead) %>% 
  ggpairs(diag = "blank",
          axisLabels = "internal",
          upper = list(continuous = wrap("cor", method = "kendall", stars = FALSE))) +
  theme(panel.grid.major = element_blank())  +
  labs(y = "Women\n")

library(cowplot)

plot_grid(
  ggmatrix_gtable(mcor),
  ggmatrix_gtable(wcor),
  nrow = 2
)
```

@@

wf_gg <- drop_rerank(wf) %>% 
  mutate(last = fct_reorder(last, -rank),
         rank = as.factor(rank),
         rank_change = ifelse(
           rank_drop %in% c(0, 1, 4, 6) |
             rank_drop == 2 & rank %in% 1:3 |
              rank_drop == 3 & rank %in% c(1, 4, 5) |
                rank_drop == 5 & rank %in% c(1, 5),
           "no", "yes")) %>%
  ggplot(aes(x = last, y = total, fill = rank_change)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = rank), hjust = -0.2, size = 3) +
  coord_flip() +
  geom_rect(aes(xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf),
            data = ~ filter(., rank_drop %in% c(2, 3, 5)), 
            color = "black", size = 1.5, fill = NA, inherit.aes = FALSE) +
  facet_wrap(~ rank_drop, nrow = 2, ncol = 4) +
  expand_limits(y = 62, x = 0:7) +
  scale_fill_manual(values = c("grey", "red")) +
  labs(y = "\nScore", x = "Women") +
  theme(axis.ticks = element_blank(),
        panel.grid.minor = element_blank())
```

@@

# Conclusion and Discussion

Overall

### 2021-07-30 manuscript/Rmd/draft.tex
Add bib file
Bed time

In 2016, the International Olympic Committee announced the addition of
five new sports to the 2020 Summer Olympics in Tokyo, Japan, which would
then reschedule for 2021 due to the impact of the COVID-19 global
pandemic. The five new features to Tokyo 2020's competitions program
include baseball/softball, karate, skateboard, sports climbing, and
surfing. One of the new sports, sport climbing, got our attention,
specifically because of its unique scoring system and the fact that only
one set of medals is awarded for each gender.

@@

At the 2020 Summer Olympics, both men's and women's sport climbing
competitions begin with 20 climbers who has previously qualified for the
Olympics from qualifying events held in 2019 and 2020. All 20 athletes
compete in each of the three disciplines in the qualification round, and
their performances in each concentration are ranked from 1 to 20. A
competitor's total score is then computed as the product of their ranks
in the three events and the lower product is better; specifically,
\begin{equation}
Score_i = R^S_i\times R^B_i\times R^L_i,
\end{equation} where \(R^S_i\), \(R^B_i\), and \(R^L_i\) are the ranks
of the \(i\)-th competitor in speed climbing, bouldering, and lead
climbing, respectively.

The top 8 finishers in the qualification round advance to the finals
where they once again compete in all three of speed climbing,
bouldering, and lead climbing. The total score for each climber in the
final stage is determined by multiplying their ranks in each discipline,
similar to the qualification round. This implies that the climber with
the lowest product of ranks in the final wins the gold medal. This type
of scoring system heavily rewards high finishes and relatively ignores
poor finishes. For instance, if climber A finished 1st, 20th, and 20th
and climber B finished 10th, 10th, and 10th, climber B would have a
score of 1000 whereas climber A would have a much better score of 400,
despite finishing last in 2 out of 3 of the events.

Heavily criticized

@@

The manuscript is outlined as follows. We first begin with some
descriptions of the data and methods in Section 2. Our analyses and
results are then presented in Sections 3 and 4. Finally, in Section 5,
we summarize up our main findings and provide a discussion to close out
the paper.

@@

Our first sets of data come from a simulation study that we conducted,
with the purpose of examining the rankings and scoring for climbers in
both qualification and final rounds. For each round, we performed 10000
simulations, and this was accomplished by randomly assigning the ranks
of each event to every participant, with the assumption that the ranks
are uniformly distributed. After the completion of the simulations, we
calculated the total scores for every simulated round, as well as the
final standings for the climbing athletes. The simulation results allow
to answer questions about various topics, including the distributions of
scores for qualifying and final rounds, and the probabilities of
advancing to the finals or winning a medal, given certain conditions.

Table 1 and Figure 1 are numerical and visual summaries of climbing
total score obtained from our simulated data for the qualification and
final rounds.

@@

For our analysis on the relationship between the rankings of the events
and the final result, we used data from the 2018 Youth Olympics Women's
Qualification. Figure 2 is a scatterplot and correlation matrix between
the ranks of the individual events and the final standings, with
Kendall's Tau (Kendall Rank Correlation Coefficient) as our measure of
ordinal association between the quantities. It is evidently clear that
there exists a strong and positive correlation between the ranks of
bouldering and lead climbing, and as a results, the standings of these
two events are highly correlated with the final rankings. On the other
hand, the correlation with the final rank is not as strong for speed
climbing. Thus, speed climbers are facing a huge disadvantage in this
scoring system, compared to those that are specialized in the other two
concentrations.

\begin{figure}
\centering
\includegraphics{draft_files/figure-latex/unnamed-chunk-9-1.pdf}
\caption{Kendall's rank correlations - 2018 World Championship, Women's
Qualification}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{draft_files/figure-latex/unnamed-chunk-12-1.pdf}
\caption{This figure illustrates the changes to the final rankings of
the 2018 Youth Olympics Men's and Women's Finals when we leave out one
climber. For each gender competition, each panel represents the rank of
the drop-out athlete, with 0 being the original final results. Each case
with a change in rank orderings is highlighted by a black panel border,
and any player with a rank change is represented by a red-filled bar}
\end{figure}

\bibliographystyle{agsm}
\bibliography{}

>>>>

In 2016, the International Olympic Committee announced the addition of
five new sports to the 2020 Summer Olympics in Tokyo, Japan, which would
then reschedule for 2021 due to the impact of the COVID-19 global
pandemic. The five new features to Tokyo 2020's competitions program
include baseball/softball, karate, skateboard, sports climbing, and
surfing \citep{ioc2016}. One of the new sports, sport climbing, got our
attention, specifically because of its unique scoring system and the
fact that only one set of medals is awarded for each gender.

@@

The decision to combine the three climbing events and only award one set
of medals for both men's and women's events in the Olympics has received
a large amount of criticism from climbing athletes all over the world.
In a series of interviews conducted by Climbing Magazine in 2016, a
number of climbers shared their thoughts and concerns about the new
Olympics climbing format. Climber Lynn Hill compared the idea of
combining speed climbing, bouldering, and lead climbing to ``asking a
middle distance runner to compete in the sprint''. She then added
``Speed climbing is a sport within our sport''. Other climbers also hold
the same opinion as Hill regarding speed climbing, using words and
phrases like ``bogus'', ``a bummer'', ``less than ideal'', ``not in
support'', and ``cheesy and unfair'' to describe the new combined
competition format. Courtney Woods stated ``Speed climbers will have the
biggest disadvantage because their realm isn't based on difficult
movements''. Mike Doyle believed ``Honestly, the people that will suffer
the most are the ones that focus only on speed climbing. Those
skills/abilities don't transfer as well to the other disciplines''. Most
climbers also expressed their hope for a change in the format in future
tournaments, with some calling for each discipline to have its own set
of medals. See \citet{blanchard2016} for more information on these
interviews.

In terms of participation and scoring, both sport climbing competitions
for male and female at the 2020 Summer Olympics begin with 20 climbers
who has previously qualified for the Olympics from qualifying events
held in 2019 and 2020. All 20 athletes compete in each of the three
disciplines in the qualification round, and their performances in each
concentration are ranked from 1 to 20. A competitor's combined score is
computed as the product of their ranks in the three events;
specifically, \begin{equation}
Score_i = R^S_i\times R^B_i\times R^L_i,
\end{equation} where \(R^S_i\), \(R^B_i\), and \(R^L_i\) are the ranks
of the \(i\)-th competitor in speed climbing, bouldering, and lead
climbing, respectively.

The top 8 lowest-scoring climbers in the qualification round advance to
the finals where they once again compete in all three of speed climbing,
bouldering, and lead climbing. The total score for each person in the
final stage is determined by multiplying their ranks in each discipline,
similar to the qualification round. This implies that the climber with
the lowest product of ranks in the final wins the gold medal. This type
of scoring system heavily rewards high finishes and relatively ignores
poor finishes. For instance, if climber A finished 1st, 20th, and 20th
and climber B finished 10th, 10th, and 10th, climber B would have a
score of 1000 whereas climber A would have a much better score of 400,
despite finishing last in 2 out of 3 of the events.

@@

\url{https://www.climbing.com/competition/olympics/a-guide-to-the-olympic-climbing-format/}

Soccer points: \citet{haugen2008}

Cross country scoring: \citet{hammond2007}, \citet{boudreau2018}

Decathlon scoring: \citet{westera2006}

The manuscript is outlined as follows. We first begin with some
descriptions of the data and methods in Section \ref{data-and-methods}.
Our analyses and results are then presented in Sections
\ref{simulations} and \ref{data-analysis}. Finally, in Section
\ref{conclusion-and-discussion}, we summarize up our main findings and
provide a discussion to close out the paper.

@@

Our first sets of data come from a simulation study that we conducted,
with the purpose of examining the rankings and scoring for climbers in
both qualification and final rounds. For each round, we performed 10000
simulations, and this was accomplished by randomly assigning the ranks
of each event to every participant, with the assumption that the ranks
are uniformly distributed. After the completion of the simulations, we
calculated the total scores for every simulated round, as well as the
final standings for the climbing athletes. The simulation results allow
to answer questions about various topics, including the distributions of
scores for qualifying and final rounds, and the probabilities of
advancing to the finals or winning a medal, given certain conditions.
First of all, Table 1 and Figure 1 are numerical and visual summaries of
climbing total score obtained from our simulated data for the
qualification and final rounds.

@@

For our analysis on the relationship between the rankings of the events
and the final result, we used data from the 2018 Youth Olympics Women's
Qualification. Figure 2 is a scatterplot and correlation matrix between
the ranks of the individual events and the final standings, with
Kendall's Tau \citep{kendall1938} as our measure of ordinal association
between the quantities. It is evidently clear that there exists a strong
and positive correlation between the ranks of bouldering and lead
climbing, and as a results, the standings of these two events are highly
correlated with the final rankings. On the other hand, the correlation
with the final rank is not as strong for speed climbing. Thus, speed
climbers are facing a huge disadvantage in this scoring system, compared
to those that are specialized in the other two concentrations.

\begin{figure}[H]
\centering
\includegraphics{draft_files/figure-latex/unnamed-chunk-9-1.pdf}
\caption{Kendall's rank correlations - 2018 World Championship, Men's
and Women's Qualifications}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics{draft_files/figure-latex/unnamed-chunk-12-1.pdf}
\caption{This figure illustrates the changes to the final rankings of
the 2018 Youth Olympics Men's and Women's Finals when we leave out one
climber. For each gender competition, each panel represents the rank of
the drop-out athlete, with 0 being the original final results. Each case
with a change in rank orderings is highlighted by a black panel border,
and any player with a rank change is represented by a red-filled bar}
\end{figure}

\hypertarget{conclusion-and-discussion}{%
\section{Conclusion and Discussion}\label{conclusion-and-discussion}}

Overall

\bibliographystyle{agsm}
\bibliography{references.bib}
